Expression QTL hotspot dissection with mediation analysis and pleiotropy testing

### 2019-07-28 analysis/paper/hotspots-paper.Rmd
added files for dift sections, to be assembled in hotspots-paper.Rmd

---
title: "Expression QTL hotspot dissection with mediation analysis and pleiotropy testing"
author: "Frederick J. Boehm, Brian S. Yandell, and Karl W. Broman"
date: "7/27/2019"
bibliography: research.bib
output: 
  - bookdown::html_document2
  - github_document:
      toc: true
      toc_depth: 2
---

# Introduction

Genetics studies in model organisms like mice can identify
genomic regions that affect complex traits, such as systolic blood pressure and body weight [@sax1923association;@soller1976power;@lander1989mapping;@broman2009guide;@jansen2007quantitative]. 
These genomic regions are called "quantitative trait loci" or "QTL".
A genome-wide QTL "scan" reveals
associations between genotypes and phenotypes by considering each position along the genome,
one at a time, as a candidate QTL for the trait of interest.
A region with strong evidence of association with a complex trait, then, defines a QTL (for that trait).
Because nearby markers have correlated
genotypes, a QTL in a two-parent cross often spans multiple megabases in length
and may contain more than a hundred genes.
Identification of a causal gene (for a given complex trait) from among those genes near the QTL is
challenging and may require costly and time-consuming experiments.
The growing need for greater QTL mapping resolution fueled development over the
last two decades of model organism multiparental populations for high-resolution QTL mapping [@de2017back;@churchill2004collaborative;@svenson2012high;@huang2012multiparent;@shivakumar2018soybean;@huang2011analysis;@kover2009multiparent;@tisne2017identification;@stanley2017genetic].
With experimentalists now measuring tens of thousands of biomolecular traits
in multiparental populations, the systems genetics community needs multivariate
statistical tools to fully examine the large volumes of data [@keller2018genetic;@chick2016defining].
Identifying loci that affect multiple traits can aid in identifying biomolecular interactions and in clarifying complex trait genetic architecture. A QTL that affects more than one trait is called a "pleiotropic" QTL.
A test of pleiotropy vs. separate QTL is one multivariate statistical tool that will inform
complex trait genetics by enabling researchers to identify the number of
unique QTL in a genomic region of interest.

@@

\section{Methods}

@keller2018genetic identified five expression trait hotspots in their
study of pancreatic islet gene expression in 378 Diversity Outbred mice. They
defined a hotspot to be a 4-Mb region that affects at least 100 nonlocal expression
traits.

For each hotspot, we identified a set of local expression traits and a
set of nonlocal expression traits. All traits demonstrated LOD peaks that met
the genome-wide significance threshold, 7.18 [@keller2018genetic].
The nonlocal expression traits arose from genes
on other chromosomes, while the local traits reflected expression levels for
genes on the same chromosome as the hotspot and within 4 Mb of the hotspot
center. We ignored transcripts for genes on the hotspot chromosome but not within 4 Mb of
the hotspot center.

## Pleiotropy testing

### Two-dimensional, two-QTL scans

For each hotspot, we tested all pairs of traits, regardless of whether the traits were local or nonlocal. We performed a
two-dimensional, two-QTL scan over a genomic region that included the entirety of the
hotspot. We fitted bivariate linear mixed effects models at each ordered pair of markers in a two-dimensional grid [@boehm2019testing].

$$
vec(Y) = X vec(B) + vec(G) + vec(E)
\label{eq:mvlmm}
$$

$$
G \sim MN_{n x 2}(0, K, V_g)
$$

$$
E \sim MN_{n x 2}(0, I_n, V_e)
$$

and $G$ and $E$ are independent.

$$
\hat B = (X^T\hat \Sigma^{-1}X)^{-1}X^T\hat\Sigma^{-1}Y
\label{eq:Bhat}
$$

$$
\hat \Sigma = \hat V_g \otimes \hat K + \hat V_e \otimes I_n
\label{eq:Sigmahat}
$$

$$
l_{10}(\hat B, \hat \Sigma) = - \frac{1}{2}\log_{10}\left((Y - X\hat B)^T\hat \Sigma^{-1}(Y - X\hat B)\right) - n\log_{10}(2\pi) - \log_{10}|\hat \Sigma|
\label{eq:loglik}
$$

For $n$ mice, $\Sigma$ is a $2n$ by $2n$ covariance matrix, while $V_g$ and $V_e$ are $2$ by $2$ covariance matrices. Estimates $\hat V_g$ and $\hat V_e$ are obtained via an expectation-maximization algorithm [@zhou2014efficient;@gemma2].

### Calculating likelihood ratio test statistics}
From the two-dimensional scans, we calculated pleiotropy likelihood ratio test statistics for each
pair of traits [@jiang1995multiple;@boehm2019testing;@boehm2019thesis]. Because of the required
computing time for bootstrap analyses, we didn't obtain p-values for the pleiotropy tests. Instead,
we worked directly with the test statistic values.

$$
LOD = \log_{10} \left(\frac{\max_{\lambda_1, \lambda_2}L(B, \Sigma, \lambda_1, \lambda_2)}{\max_{\lambda}L(B, \Sigma, \lambda, \lambda)} \right)
\label{eq:lrt}
$$

$$
ll(\lambda_1, \lambda_2)
$$

## Mediation analysis

A mediation analysis in systems genetics begins with an identified QTL-expression trait association. In our case, we choose a QTL that affects a nonlocal trait.

For each hotspot and each local trait - nonlocal trait pair, we also performed a
mediation analysis to assess the extent which the local trait mediates the QTL effect on the nonlocal trait. Each mediation analysis involved fitting four linear models (Equations~\ref{model1}, \ref{model2}, \ref{model3}, \ref{model4}) [@chick2016defining].

$$
Y = \alpha 1 + WC + E
\label{model1}
$$

$$
Y = XB + WC + E
\label{model2}
$$

$$
Y = \alpha 1 + WC + M\beta + E
\label{model3}
$$

$$
Y = XB + WC + M\beta + E
\label{model4}
$$

In all mediation analyses, the nonlocal transcripts are the targets, while the local
transcripts serve as candidate mediators of the QTL-target relationship.
In Equations~\ref{model1} through \ref{model4}, $X$ is a matrix of founder allele dosages,
$B$ is a matrix
of founder allele effects, $W$ is a matrix of covariates, $C$ is a matrix of covariate
effects, $M$ is a vector of values for a single candidate mediator, $\beta$ is the effect
of the mediator on the complex trait, 1 is a vector with all entries set to 1, $\alpha$
is the mean trait value, and $E$ is a vector of normally distributed random errors.

## Visualizations

We created scatter plots with the `ggplot2` package [@ggplot2] for the R
statistical computing environment [@r]. We initially plotted all local-nonlocal pairs for each hotspot. We highlighted in blue those points that correspond to pairs involving the putative mediator per [@keller2018genetic].

We created heatmaps to examine patterns among pleiotropy test results in each hotspot.
We represented the pleiotropy test statistics as a symmetric matrix, where each row was one expression
trait and each column was an expression trait. Each cell, then, contained the pleiotropy test statistic
from the analysis involving the row's expression trait and the column's expression trait.
To create a heatmap, we first performed two hierarchical clusterings, one on the submatrix involving only
nonlocal expression traits and a second on the submatrix involving only local expression traits. We
calculated the Euclidean distances between all trait pairs before clustering. We ordered the rows and
columns according to optimal leaf ordering. We then combined the
local and nonlocal traits into a single heatmap. To achieve greater coloring contrasts in the heatmaps, we
set the color spectrum limits at 0 (dark blue) and 5 (yellow). Test statistics above 5 are represented by
yellow.

@@

```{r more_setup-2}
hot_lower <- hot_mid - 2
hot_upper <- hot_mid + 2
fp <- paste0("Chr", hot_chr, "-")
knitr::opts_chunk$set(fig.path = here("analysis", "figures", fp))
```

```{r readRDS-2}
pleio_mixed_list <- readRDS(here("analysis", "data", "derived_data", paste0("Chr", hot_chr, "_pleio.rds")))
pleio_local_list <- readRDS(here("analysis", "data", "derived_data", paste0("Chr", hot_chr, "_pleio_local.rds")))
pleio_nonlocal_list <- readRDS(here("analysis", "data", "derived_data", paste0("Chr", hot_chr, "_pleio_nonlocal.rds")))
med_list <- readRDS(here("analysis", "data", "derived_data", paste0("Chr", hot_chr, "_mediations.rds")))
```

```{r load_attie, eval=!file.exists(here("analysis", "data", "derived_data", "expr.rds"))}
download.file("https://datadryad.org/bitstream/handle/10255/dryad.166774/Attie_DO378_eQTL_viewer_v1.Rdata?sequence=2",
              destfile = here("analysis", "data", "raw_data", "Attie_DO378_eQTL_viewer_v1.Rdata")
              )
load(here("analysis", "data", "raw_data", "Attie_DO378_eQTL_viewer_v1.Rdata"))
saveRDS(dataset.islet.rnaseq$lod.peaks, here("analysis", "data", "derived_data", "lod_peaks.rds"))
saveRDS(dataset.islet.rnaseq$annots, here("analysis", "data", "derived_data", "annots.rds"))
saveRDS(dataset.islet.rnaseq$expr, here("analysis", "data", "derived_data", "expr.rds"))
```

```{r read_attie_files}
lod_peaks <- readRDS(here("analysis", "data", "derived_data", "lod_peaks.rds"))
annots <- readRDS(here("analysis", "data", "derived_data", "annots.rds"))
expr <- readRDS(here("analysis", "data", "derived_data", "expr.rds"))
```

```{r make_local_annot-2}
local_annot <- lod_peaks %>%
  filter(chrom == hot_chr, pos <= hot_upper, pos >= hot_lower) %>%
  left_join(y = annots, by = c("annot.id" = "gene_id")) %>%
  filter(!duplicated(annot.id)) %>%
  filter(lod >= ult) %>%
  mutate(local_tx = chr == hot_chr) %>%
  filter(local_tx)
```

```{r make_nonlocal_annot-2}
nonlocal_annot <- lod_peaks %>%
  filter(chrom == hot_chr, pos <= hot_upper, pos >= hot_lower) %>%
  inner_join(y = annots, by = c("annot.id" = "gene_id")) %>%
  filter(!duplicated(annot.id)) %>%
  filter(lod >= 7.18) %>%
  mutate(local_tx = chr == hot_chr) %>%
  filter(!local_tx) %>%
  filter(!is.na(hotspot))
```

```{r make_local_annot2-2}
local_annot2 <- local_annot %>%
  filter(middle <= hot_upper + 5, middle >= hot_lower - 5) %>%
  arrange(desc(lod))
```

```{r subset_expr_matrix-2}
hot_local <- expr[, colnames(expr) %in% local_annot2$annot.id]
hot_nonlocal <- expr[, colnames(expr) %in% nonlocal_annot$annot.id]
nnonlocal <- nrow(nonlocal_annot)
nlocal <- nrow(local_annot2)
```

```{r tabulate-2, message=FALSE, warning = FALSE}
med_tib <- med_list %>%
  purrr::map(bind_rows) %>%
  bind_rows() %>%
  as_tibble() %>%
  rename(lod_without_med = lod_no_med, lod_with_med = lod_med) %>%
  mutate(lod_diff = lod_without_med - lod_with_med, lod_diff_proportion = lod_diff / lod_without_med) %>%
  left_join(local_annot2, by = c("local_gene_id" = "annot.id")) %>%
  left_join(nonlocal_annot, by = c("nonlocal_gene_id" = "annot.id")) %>%
  dplyr::select(symbol.x, symbol.y, lod_without_med, lod_with_med, lod_diff, lod_diff_proportion, local_gene_id, nonlocal_gene_id) %>%
  rename(local_symbol = symbol.x, nonlocal_symbol = symbol.y) %>%
  filter(!is.na(local_symbol))
```

```{r make_pleio_tib-2}
hot_annot <- local_annot2 %>%
  bind_rows(nonlocal_annot) %>%
  dplyr::select(annot.id, symbol)
pleio_tib <- pleio_mixed_list %>%
  purrr::map(.f = function(x) rename(x, gene1_id = local_gene_id, gene2_id = nonlocal_gene_id)) %>%
  c(pleio_local_list, pleio_nonlocal_list) %>%
  bind_rows() %>%
  rename(pleiotropy_lod = lod) %>%
  inner_join(hot_annot, by = c("gene1_id" = "annot.id")) %>%
  rename(gene1_symbol = symbol) %>%
  inner_join(hot_annot, by = c("gene2_id" = "annot.id")) %>%
  rename(gene2_symbol = symbol) %>%
  dplyr::select(gene1_symbol, gene2_symbol, pleiotropy_lod)
```

```{r make_pm-2}
pm <- pleio_tib %>%
  tibble_to_matrix(symmetric = TRUE)
```

```{r scatter-2}
p <- pleio_mixed_list %>%
  bind_rows() %>%
  rename(pleiotropy_lod = lod) %>%
  inner_join(med_tib, by = c("local_gene_id", "nonlocal_gene_id")) %>%
  dplyr::select(local_symbol, nonlocal_symbol, pleiotropy_lod, lod_diff, lod_diff_proportion, lod_without_med, lod_with_med, filename, nonlocal_gene_id, local_gene_id) %>%
  ggplot() + geom_point(mapping = aes(x = pleiotropy_lod, y = lod_diff, colour = local_symbol == keller_mediator, nonlocal = nonlocal_symbol), size = 0.05) + theme(legend.position = "none") + xlab("Pleiotropy LOD") + ylab("Mediation LOD difference")
fnsvg <- paste0("Chr", hot_chr, "_scatter.svg")
fnpdf <- paste0("Chr", hot_chr, "_scatter.pdf")
ggsave(plot = p, filename = here("analysis", "figures", fnsvg))
ggsave(plot = p, filename = here("analysis", "figures", fnpdf))
p
```

```{r scatter_by_local-2}
foo <- pleio_mixed_list %>%
  bind_rows() %>%
  rename(pleiotropy_lod = lod) %>%
  inner_join(med_tib, by = c("local_gene_id", "nonlocal_gene_id")) %>%
  dplyr::select(local_symbol, nonlocal_symbol, pleiotropy_lod, lod_diff, lod_diff_proportion, lod_without_med, lod_with_med, filename, nonlocal_gene_id, local_gene_id)
foo %>%
  ggplot() +
  geom_point(mapping = aes(x = pleiotropy_lod, y = lod_diff, colour = nonlocal_symbol, nonlocal = nonlocal_symbol), size = 0.1) +
  facet_wrap(. ~ local_symbol, nrow = ceiling(nlocal / 2), ncol = 2) +
  broman::karl_theme(strip.placement = "inside") +
  theme(legend.position = "none") +
  theme(
    strip.background = element_blank(),
    strip.text.x = element_blank()
) +
  xlab("Pleiotropy LOD") +
  ylab("Mediation LOD difference") +
  geom_text(aes(x = 1, y = 1, label = local_symbol), inherit.aes=FALSE)

fnsvg <- paste0("Chr", hot_chr, "_scatter-panel.svg")
fnpdf <- paste0("Chr", hot_chr, "_scatter-panel.pdf")
ggsave(filename = here("analysis", "figures", fnsvg), height = ceiling(nlocal / 2))
ggsave(filename = here("analysis", "figures", fnpdf), height = ceiling(nlocal / 2))

# put in multiple figures, if there are many local traits
n_figs <- 1 + nlocal %/% 18 # 18 is the number of panels per page in pdf
n_last <- nlocal %% 18 # number of panels on last figure
max_pleio_lod <- max(foo$pleiotropy_lod)
max_med_lod_difference <- max(foo$lod_diff)
min_med_lod_difference <- min(foo$lod_diff)

for (i in 1:n_figs){
  n_panels <- 18 * (i < n_figs) + n_last * (i == n_figs)
  foo %>%
    filter(local_symbol %in% local_annot2$symbol[((i - 1) * 18 + 1): (18 * i)]) %>%
    ggplot() +
      geom_point(mapping = aes(x = pleiotropy_lod, y = lod_diff, colour = nonlocal_symbol, nonlocal = nonlocal_symbol), size = 0.1) +
      facet_wrap(. ~ local_symbol, nrow = ceiling(n_panels / 2), ncol = 2) +
      broman::karl_theme() +
      theme(legend.position = "none") +
      theme(
        strip.background = element_blank(),
        strip.text.x = element_blank()
    ) +
      xlab("Pleiotropy LOD") + xlim(c(0, max_pleio_lod)) + ylim(c(min_med_lod_difference, max_med_lod_difference)) +
      ylab("Mediation LOD difference") +
      geom_text(size = 2, aes(x = max_pleio_lod / 2, y = 2 * max_med_lod_difference / 3, label = local_symbol), inherit.aes=FALSE)
  fnsvg <- paste0("Chr", hot_chr, "_scatter-panel", i, ".svg")
  fnpdf <- paste0("Chr", hot_chr, "_scatter-panel", i, ".pdf")
  ggsave(filename = here("analysis", "figures", fnsvg), height = ceiling(n_panels / 2))
  ggsave(filename = here("analysis", "figures", fnpdf), height = ceiling(n_panels / 2))
}
```

```{r make_scols-2}
scols1 <- 1L + (rownames(pm) %in% local_annot2$symbol)
scols2 <- rownames(pm) %in% (med_tib %>% filter(local_symbol == keller_mediator, lod_diff > 1.5))$nonlocal_symbol
```

```{r heat8-2}
# order the nonlocal traits first
## indicator for being a nonlocal trait
indic_nonlocal <- rownames(pm) %in% nonlocal_annot$symbol
# define ord_nonlocal
ord_nonlocal <- pm[indic_nonlocal, indic_nonlocal] %>%
  dist() %>%
  hclust() %>%
  as.dendrogram() %>%
  dendextend::seriate_dendrogram(dist(pm[indic_nonlocal, indic_nonlocal])) %>%
  order.dendrogram()
rn_nonlocal <- rownames(pm[indic_nonlocal, indic_nonlocal][ord_nonlocal, ord_nonlocal])
# define ord_local
ord_local<- pm[!indic_nonlocal, !indic_nonlocal] %>%
  dist() %>%
  hclust() %>%
  as.dendrogram() %>%
  dendextend::seriate_dendrogram(dist(pm[!indic_nonlocal, !indic_nonlocal])) %>%
  order.dendrogram()
rn_local <- rownames(pm[!indic_nonlocal, !indic_nonlocal][ord_local, ord_local])
rn <- c(rn_nonlocal, rn_local)
```

```{r make_scols-extra-2}
scols_extra <- purrr::map(.x = local_annot2$symbol,
                          .f = function(x) {rownames(pm) %in% (med_tib %>% filter(local_symbol == x, lod_diff > 1.5))$nonlocal_symbol}
) %>%
  bind_cols() %>%
  (function(x){colnames(x) <- local_annot2$symbol; x}) %>%
  mutate(local = scols1)
```

```{r count_med}
n_med_tr <- 4 # arbitrary threshold
n_med_per_local <- med_tib %>%
  group_by(local_symbol) %>%
  filter(lod_diff > 1.5) %>%
  count() %>%
  filter(n >= n_med_tr)
n_med_per_local %>%
  arrange(desc(n)) %>%
  knitr::kable()

scols_pre <- scols_extra %>%
  dplyr::select(n_med_per_local$local_symbol)
# get ordering for columns of scols_pre
col_order <- scols_pre %>%
  t() %>%
  dist() %>%
  hclust() %>%
  as.dendrogram() %>%
  dendextend::seriate_dendrogram(dist(t(scols_pre))) %>%
  order.dendrogram()
scols <- scols_pre[ , col_order] %>%
  mutate(local = scols1)
fnhtml <- here("analysis", "figures", paste0("Chr", hot_chr, "hm.html"))
```

```{r heat9-2, eval = !file.exists(fnhtml)}
ord <- match(rn, rownames(pm))

p <- pm[ord, ord] %>%
  apply(FUN = trunc2, MARGIN = 2, threshold = 5) %>%
  heatmaply(Rowv = FALSE,
            Colv = FALSE,
            cexRow = 0.2,
            cexCol = 0.2,
            symm = TRUE,
            row_side_colors = scols[ord, ],
            col_side_colors = tibble(local = scols1[ord]),
            limits = c(0, 5)
            ) %>%
  hide_guides()
fnpdf <- here("analysis", "figures", paste0("Chr", hot_chr, "hm.pdf"))
htmlwidgets::saveWidget(widget = p,
                        file = fnhtml
                        )
webshot::webshot(url = fnhtml,
                 file = fnpdf,
                 cliprect = "viewport"
                 )
```

```{r heat10-2}
webshot::webshot(fnhtml)
```

```{r index-v-peak-position-2}
tibble(symbol = rownames(pm[indic_nonlocal, indic_nonlocal])[ord_nonlocal], index = 1:length(symbol)) %>%
  left_join(nonlocal_annot, by = "symbol") %>%
  ggplot() + geom_point(aes(y = pos, x = index))
```

@@

```{r more_setup-5, ref.label = "more_setup-2"}
```

```{r readRDS-5, ref.label="readRDS-2"}
```

```{r make_local_annot-5, ref.label="make_local_annot-2"}
```

```{r make_nonlocal_annot-5, ref.label = "make_nonlocal_annot-2"}
```

```{r make_local_annot2-5, ref.label = "make_local_annot2-2"}
```

```{r subset_expr_matrix-5, ref.label = "subset_expr_matrix-2"}
```

```{r tabulate-5, ref.label = "tabulate-2", message=FALSE, warning = FALSE}
```

```{r make_pleio_tib-5, ref.label="make_pleio_tib-2"}
```

```{r make_pm-5, ref.label = "make_pm-2"}
```

```{r scatter-5, ref.label="scatter-2"}
```

```{r scatter_by_local-5, ref.label = "scatter_by_local-2"}
```

```{r make_scols-5, ref.label="make_scols-2"}
```

```{r heat8-5, ref.label="heat8-2"}
```

```{r make_scols-extra-5, ref.label = "make_scols-extra-2"}
```

```{r count_med-5, ref.label = "count_med"}
```

```{r heat9-5, ref.label = "heat9-2", eval = !file.exists(fnhtml)}
```

```{r heat10-5, ref.label="heat10-2"}
```

```{r index-v-peak-position-5, ref.label = "index-v-peak-position-2"}
```

@@

```{r more_setup-7, ref.label = "more_setup-2"}
```

```{r readRDS-7, ref.label="readRDS-2"}
```

```{r make_local_annot-7, ref.label="make_local_annot-2"}
```

```{r make_nonlocal_annot-7, ref.label = "make_nonlocal_annot-2"}
```

```{r make_local_annot2-7, ref.label = "make_local_annot2-2"}
```

```{r subset_expr_matrix-7, ref.label = "subset_expr_matrix-2"}
```

```{r tabulate-7, ref.label = "tabulate-2", message=FALSE, warning = FALSE}
```

```{r make_pleio_tib-7, ref.label="make_pleio_tib-2"}
```

```{r make_pm-7, ref.label = "make_pm-2"}
```

```{r scatter-7, ref.label="scatter-2"}
```

```{r scatter_by_local-7, ref.label = "scatter_by_local-2"}
```

```{r make_scols-7, ref.label="make_scols-2"}
```

```{r heat8-7, ref.label="heat8-2"}
```

```{r make_scols-extra-7, ref.label = "make_scols-extra-2"}
```

```{r count_med-7, ref.label = "count_med"}
```

```{r heat9-7, ref.label = "heat9-2", eval = !file.exists(fnhtml)}
```

```{r heat10-7, ref.label="heat10-2"}
```

```{r index-v-peak-position-7, ref.label = "index-v-peak-position-2"}
```

@@

```{r more_setup-11, ref.label = "more_setup-2"}
```

```{r readRDS-11, ref.label="readRDS-2"}
```

```{r make_local_annot-11, ref.label="make_local_annot-2"}
```

```{r make_nonlocal_annot-11, ref.label = "make_nonlocal_annot-2"}
```

```{r make_local_annot2-11, ref.label = "make_local_annot2-2"}
```

```{r subset_expr_matrix-11, ref.label = "subset_expr_matrix-2"}
```

```{r tabulate-11, ref.label = "tabulate-2", message=FALSE, warning = FALSE}
```

```{r make_pleio_tib-11, ref.label="make_pleio_tib-2"}
```

```{r make_pm-11, ref.label = "make_pm-2"}
```

```{r scatter-11, ref.label="scatter-2"}
```

```{r scatter_by_local-11, ref.label = "scatter_by_local-2"}
```

```{r make_scols-11, ref.label="make_scols-2"}
```

```{r heat8-11, ref.label="heat8-2"}
```

```{r make_scols_extra-11, ref.label = "make_scols-extra-2"}
```

```{r count_med-11, ref.label = "count_med"}
```

```{r heat9-11, ref.label = "heat9-2", eval = !file.exists(fnhtml)}
```

```{r heat10-11, ref.label="heat10-2"}
```

```{r index-v-peak-position-11, ref.label = "index-v-peak-position-2"}
```

@@

```{r more_setup-13, ref.label = "more_setup-2"}
```

```{r readRDS-13, ref.label="readRDS-2"}
```

```{r make_local_annot-13, ref.label="make_local_annot-2"}
```

```{r make_nonlocal_annot-13, ref.label = "make_nonlocal_annot-2"}
```

```{r make_local_annot2-13, ref.label = "make_local_annot2-2"}
```

```{r subset_expr_matrix-13, ref.label = "subset_expr_matrix-2"}
```

```{r tabulate-13, ref.label = "tabulate-2", message=FALSE, warning = FALSE}
```

```{r make_pleio_tib-13, ref.label="make_pleio_tib-2"}
```

```{r make_pm-13, ref.label = "make_pm-2"}
```

```{r scatter-13, ref.label="scatter-2"}
```

```{r scatter_by_local-13, ref.label = "scatter_by_local-2"}
```

```{r make_scols-13, ref.label="make_scols-2"}
```

```{r heat8-13, ref.label="heat8-2"}
```

```{r make_scols-extra-13, ref.label = "make_scols-extra-2"}
```

```{r count_med-13, ref.label = "count_med"}
```

```{r heat9-13, ref.label = "heat9-2", eval = !file.exists(fnhtml)}
```

```{r heat10-13, ref.label="heat10-2"}
```

```{r index-v-peak-position-13, ref.label = "index-v-peak-position-2"}
```

>>>>

---
title: "Expression QTL hotspot dissection with mediation analysis and pleiotropy testing"
author: "Frederick J. Boehm, Brian S. Yandell, and Karl W. Broman"
date: "7/27/2019"
bibliography: research.bib
numberSections: true
autoEqnLabels: true
output: 
  bookdown::html_document2:
    pandoc_args: 
      - --filter
      - /usr/local/Cellar/pandoc-crossref/0.3.4.1/bin/pandoc-crossref
---

```{r child="intro.Rmd"}
```

@@

```{r child="methods.Rmd"}
```

@@

```{r child="hotspot-one-chromosome.Rmd"}
```

@@

```{r child="hotspot-one-chromosome.Rmd"}
```

@@

```{r child="hotspot-one-chromosome.Rmd"}
```

@@

```{r child="hotspot-one-chromosome.Rmd"}
```

@@

```{r child="hotspot-one-chromosome.Rmd"}
```

### 2019-07-28 analysis/paper/methods.Rmd
added files for dift sections, to be assembled in hotspots-paper.Rmd

---
title: "Expression QTL hotspot dissection with mediation analysis and pleiotropy testing"
author: "Frederick J. Boehm, Brian S. Yandell, and Karl W. Broman"
date: "7/27/2019"
bibliography: research.bib
output: 
  bookdown::html_document2:
    pandoc_args: 
      - --filter
      - /usr/local/Cellar/pandoc-crossref/0.3.4.1/bin/pandoc-crossref
---

```{r load-broman}
library(broman)
library(here)
```

@@

$$
vec(Y) = X vec(B) + vec(G) + vec(E)
$$

@@

### Calculating likelihood ratio test statistics}
From the two-dimensional scans, we calculated pleiotropy likelihood ratio test statistics for each
pair of traits [@jiang1995multiple;@boehm2019testing;@boehm2019thesis]. Because of the required
computing time for bootstrap analyses, we didn't obtain p-values for the pleiotropy tests. Instead,
we worked directly with the test statistic values.

$$
LOD = \log_{10} \left(\frac{\max_{\lambda_1, \lambda_2}L(B, \Sigma, \lambda_1, \lambda_2)}{\max_{\lambda}L(B, \Sigma, \lambda, \lambda)} \right)
$$

@@

$$
Y = \alpha 1 + WC + E
\label{model1}
$$

$$
Y = XB + WC + E
\label{model2}
$$

$$
Y = \alpha 1 + WC + M\beta + E
\label{model3}
$$

$$
Y = XB + WC + M\beta + E
\label{model4}
$$

In all mediation analyses, the nonlocal transcripts are the targets, while the local
transcripts serve as candidate mediators of the QTL-target relationship.
In Equations~\ref{model1} through \ref{model4}, $X$ is a matrix of founder allele dosages,
$B$ is a matrix
of founder allele effects, $W$ is a matrix of covariates, $C$ is a matrix of covariate
effects, $M$ is a vector of values for a single candidate mediator, $\beta$ is the effect
of the mediator on the complex trait, 1 is a vector with all entries set to 1, $\alpha$
is the mean trait value, and $E$ is a vector of normally distributed random errors.

@@

# Results

We generated scatter plots of pleiotropy test statistics vs. mediation LOD difference values for each of the five hotspots (Figures).
Each point represents a pairing of a local trait with a nonlocal trait.
The mediation LOD difference value is that observed when the local trait is considered
as a candidate mediator of the QTL-nonlocal trait association.

In all five scatter plots, we observe few
points in the upper right quadrant of the figure.
That is, few pairs have high pleiotropy test statistics and
high mediation LOD difference values.

```{r setup}
library(qtl2hotspots)
pleio_threshold <- 4
ult <- 15 # univariate lod threshold
```

```{r load_tidyverse}
library(tidyverse)
options(tibble.print_max = 50, tibble.print_min = 20)
```

```{r load_heatmaply}
library(heatmaply)
```

```{r setup_continued-2}
hot_chr <- 2
hot_mid <- 165.5
keller_mediator <- "Hnf4a"
```

```{r child="hotspot-one-chromosome.Rmd"}
```

# Discussion

Pleiotropy tests complement mediation analyses in the following manner. When considering a
collection of potential mediators for a nonlocal trait, mediation analyses clearly identify the
mediator when there is, in fact, a mediator among the candidates. However, when there is no
mediator, mediation analyses provide little information about the genetics of the complex trait.
Pleiotropy testing is particularly useful in this setting, since a pleiotropy test always gives
an answer about the number of underlying QTL. A drawback of pleiotropy testing is that the scientific question that it addresses - do two traits share a single QTL - informs, but doesn't fully resolve, the genetics of the nonlocal complex traits under study.

# Colophon

This report was generated on ` r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE, echo = TRUE, tidy.opts=list(width.cutoff=50), size = "tiny", results = "markup"}
# which R packages and versions?
devtools::session_info()
```

The current Git commit details are:

```{r git2r, cache = FALSE, echo = TRUE}
# what commit is this file at?
git2r::repository(here::here())
```

>>>>

@@

$$
vec(Y) = X vec(B) + vec(G) + vec(E)
$${eq:mvlmm}

@@

### Calculating likelihood ratio test statistics

From the two-dimensional scans, we calculated pleiotropy likelihood ratio test statistics for each
pair of traits [@jiang1995multiple;@boehm2019testing;@boehm2019thesis]. Because of the required
computing time for bootstrap analyses, we didn't obtain p-values for the pleiotropy tests. Instead,
we worked directly with the test statistic values.

$$
LOD = \log_{10} \left(\frac{\max_{\lambda_1, \lambda_2}L(B, \Sigma, \lambda_1, \lambda_2)}{\max_{\lambda}L(B, \Sigma, \lambda, \lambda)} \right)
$${eq:lod}

@@

$$
Y = \alpha 1 + WC + E
$${eq:model1}

$$
Y = XB + WC + E
$${eq:model2}

$$
Y = \alpha 1 + WC + M\beta + E
$${eq:model3}

$$
Y = XB + WC + M\beta + E
$${eq:model4}

In all of our mediation analyses, the nonlocal transcripts are the targets, while the local
transcripts serve as candidate mediators of the QTL-target relationship.
In Equations~\ref{model1} through \ref{model4}, $X$ is a matrix of founder allele dosages,
$B$ is a matrix
of founder allele effects, $W$ is a matrix of covariates, $C$ is a matrix of covariate
effects, $M$ is a vector of values for a single candidate mediator, $\beta$ is the effect
of the mediator on the complex trait, 1 is a vector with all entries set to 1, $\alpha$
is the mean trait value, and $E$ is a vector of normally distributed random errors.

@@
